---
title: "Flagler Health: Exploratory Analysis"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

library(dplyr)
library(ggplot2)
library(tidyr)

project_root <- normalizePath("..", mustWork = FALSE)
knitr::opts_knit$set(root.dir = project_root)

data_dir <- file.path(project_root, "data")
metrics_path <- file.path(data_dir, "pain_rtm_longitudinal.csv")
```

# 1. Load and Prepare Data

```{r load-data}
metrics <- read.csv(metrics_path, stringsAsFactors = FALSE)

# Standardize blank strings as missing values.
metrics[metrics == ""] <- NA

metrics <- metrics %>%
  mutate(
    date = as.Date(date),
    nprs_pain_val = as.numeric(nprs_pain_val),
    across(c(mobility_val, mood_val, sleep_val), as.numeric)
  )
```

## Missingness Summary

```{r missingness}
missingness <- tibble(
  variable = names(metrics),
  missing_pct = colMeans(is.na(metrics)) * 100
) %>%
  arrange(desc(missing_pct))

knitr::kable(
  missingness,
  digits = 1,
  col.names = c("Variable", "Missing (%)")
)
```

# 2. Longitudinal Patient Examples

The original analysis included two example patient trajectories. Patient identifiers and direct message quotations are intentionally not stored in this notebook.

```{r patient-trajectory-helper}
plot_pain_trajectory <- function(data, patient_id, outcome_label) {
  patient_pain <- data %>%
    filter(patient_id == !!patient_id, !is.na(nprs_pain_val)) %>%
    arrange(date)

  if (nrow(patient_pain) == 0) {
    stop("No pain observations found for the supplied patient ID.")
  }

  # Highlight meaningful points without exposing message text or hard-coded dates.
  highlights <- bind_rows(
    patient_pain %>% slice_min(date, n = 1, with_ties = FALSE) %>% mutate(event = "First observation"),
    patient_pain %>% slice_max(nprs_pain_val, n = 1, with_ties = FALSE) %>% mutate(event = "Peak pain"),
    patient_pain %>% slice_max(date, n = 1, with_ties = FALSE) %>% mutate(event = "Last observation")
  ) %>%
    distinct(date, nprs_pain_val, .keep_all = TRUE)

  ggplot(patient_pain, aes(x = date, y = nprs_pain_val)) +
    geom_line(linewidth = 0.9) +
    geom_point(size = 2.5) +
    geom_point(data = highlights, size = 3) +
    geom_text(
      data = highlights,
      aes(label = event),
      vjust = -0.8,
      check_overlap = TRUE
    ) +
    scale_y_continuous(limits = c(0, 10), breaks = 0:10) +
    labs(
      title = paste("Pain Score Over Time —", outcome_label),
      subtitle = "Patient identifier redacted",
      x = "Date",
      y = "Pain Score (NPRS)"
    ) +
    theme_minimal() +
    theme(panel.grid.minor = element_blank())
}
```

```{r retained-example, eval=TRUE}
retained_patient_id <- Sys.getenv("FLAGLER_RETAINED_EXAMPLE_ID")

if (nzchar(retained_patient_id)) {
  print(plot_pain_trajectory(metrics, retained_patient_id, "Patient Still Enrolled"))
} else {
  message("Retained-patient example skipped: FLAGLER_RETAINED_EXAMPLE_ID is not set.")
}
```

```{r dropout-example, eval=TRUE}
dropout_patient_id <- Sys.getenv("FLAGLER_DROPOUT_EXAMPLE_ID")

if (nzchar(dropout_patient_id)) {
  print(plot_pain_trajectory(metrics, dropout_patient_id, "Patient Dropped Off"))
} else {
  message("Dropout-patient example skipped: FLAGLER_DROPOUT_EXAMPLE_ID is not set.")
}
```

# 3. Aggregate Exploratory Analysis

## Distribution of Pain Scores

```{r pain-distribution}
ggplot(metrics, aes(x = nprs_pain_val)) +
  geom_bar() +
  scale_x_continuous(breaks = 0:10) +
  labs(
    title = "Distribution of Reported Pain Scores",
    x = "Pain Score (NPRS)",
    y = "Number of Reports"
  ) +
  theme_minimal()
```

## Engagement Summary

```{r engagement-summary}
engagement <- metrics %>%
  filter(!is.na(chat_id), !is.na(date)) %>%
  group_by(chat_id) %>%
  summarise(
    total_reports = n(),
    avg_pain = mean(nprs_pain_val, na.rm = TRUE),
    duration_days = as.numeric(max(date) - min(date)),
    .groups = "drop"
  )

engagement_overview <- engagement %>%
  summarise(
    conversations = n(),
    median_reports = median(total_reports, na.rm = TRUE),
    median_duration_days = median(duration_days, na.rm = TRUE),
    median_avg_pain = median(avg_pain, na.rm = TRUE)
  )

knitr::kable(engagement_overview, digits = 1)
```

## Sleep Quality and Pain

```{r sleep-pain}
ggplot(metrics, aes(x = sleep_val, y = nprs_pain_val)) +
  geom_jitter(alpha = 0.25, width = 0.08, height = 0.08) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Sleep Quality vs. Reported Pain",
    x = "Sleep Score",
    y = "Pain Score (NPRS)"
  ) +
  theme_minimal()
```

## Monthly Pain Trend

```{r monthly-trend}
monthly_trend <- metrics %>%
  filter(!is.na(date)) %>%
  mutate(month = format(date, "%Y-%m")) %>%
  group_by(month) %>%
  summarise(
    avg_pain = mean(nprs_pain_val, na.rm = TRUE),
    reports = sum(!is.na(nprs_pain_val)),
    .groups = "drop"
  )

ggplot(monthly_trend, aes(x = month, y = avg_pain, group = 1)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  labs(
    title = "Average Reported Pain by Month",
    x = "Month",
    y = "Mean Pain Score"
  ) +
  theme_minimal()
```
