---
title: "Flagler Health: Survival and Logistic Modeling"
output:
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)

library(dplyr)
library(tidyr)
library(knitr)
library(survival)
library(broom)
library(ggplot2)

project_root <- normalizePath("..", mustWork = FALSE)
knitr::opts_knit$set(root.dir = project_root)

data_dir <- file.path(project_root, "data")
patient_path <- file.path(data_dir, "patients_w_baseline.csv")
sentiment_path <- file.path(data_dir, "patient_daily_sentiment.csv")
```

# 1. Prepare Patient-Level Modeling Data

```{r prepare-data}
patients_raw <- read.csv(patient_path, stringsAsFactors = FALSE)

patients <- patients_raw %>%
  transmute(
    patient_id,
    days_enrolled,
    dropout = ifelse(is_enrolled, 0L, 1L),
    pain = baseline_nprs_pain_val,
    mobility = factor(baseline_mobility_val),
    mood = factor(baseline_mood_val),
    sleep = factor(baseline_sleep_val),
    procedure = factor(baseline_procedure),
    diag_chapter = factor(baseline_diag_chapter)
  ) %>%
  drop_na(
    patient_id,
    days_enrolled,
    dropout,
    pain,
    mobility,
    mood,
    sleep,
    procedure,
    diag_chapter
  ) %>%
  mutate(
    procedure_2group = if_else(
      procedure == "Evaluation and Management",
      "Evaluation and Management",
      "Other"
    ),
    procedure_3group = case_when(
      procedure == "Evaluation and Management" ~ "Evaluation and Management",
      procedure == "Surgery" ~ "Surgery",
      TRUE ~ "Other"
    ),
    diagnosis_2group = if_else(
      diag_chapter == "Diseases of the Musculoskeletal System and Connective Tissue",
      "Musculoskeletal",
      "Other"
    ),
    procedure_2group = factor(
      procedure_2group,
      levels = c("Evaluation and Management", "Other")
    ),
    procedure_3group = factor(
      procedure_3group,
      levels = c("Evaluation and Management", "Surgery", "Other")
    ),
    diagnosis_2group = factor(
      diagnosis_2group,
      levels = c("Musculoskeletal", "Other")
    )
  )

stopifnot(all(patients$dropout %in% c(0L, 1L)))
```

# 2. Procedure Group Summary

```{r procedure-summary}
procedure_summary <- patients %>%
  count(procedure, name = "count") %>%
  arrange(desc(count))

kable(procedure_summary, caption = "Baseline Procedure Groups")

procedure_2group_summary <- patients %>%
  count(procedure_2group, name = "count")

kable(
  procedure_2group_summary,
  caption = "Baseline Procedure Groups — Two-Group Version"
)
```

# 3. Kaplan-Meier Analysis

## Overall Retention Curve

```{r km-overall}
overall_km <- survfit(Surv(days_enrolled, dropout) ~ 1, data = patients)

plot(
  overall_km,
  main = "Kaplan-Meier Estimate of RTM Retention",
  xlab = "Days",
  ylab = "Proportion Enrolled"
)
```

## RTM Baseline Metrics

```{r km-baseline-metrics, fig.height=12}
par(mfrow = c(3, 1))

km_mobility <- survfit(Surv(days_enrolled, dropout) ~ mobility, data = patients)
plot(km_mobility, main = "Retention by Baseline Mobility", xlab = "Days", ylab = "Proportion Enrolled")
legend("bottomright", legend = levels(patients$mobility), lty = 1, bty = "n")

km_mood <- survfit(Surv(days_enrolled, dropout) ~ mood, data = patients)
plot(km_mood, main = "Retention by Baseline Mood", xlab = "Days", ylab = "Proportion Enrolled")
legend("bottomright", legend = levels(patients$mood), lty = 1, bty = "n")

km_sleep <- survfit(Surv(days_enrolled, dropout) ~ sleep, data = patients)
plot(km_sleep, main = "Retention by Baseline Sleep", xlab = "Days", ylab = "Proportion Enrolled")
legend("bottomright", legend = levels(patients$sleep), lty = 1, bty = "n")

par(mfrow = c(1, 1))
```

## Pain Quartiles

```{r pain-groups}
pain_breaks <- unique(
  quantile(
    patients$pain,
    probs = c(0, 0.25, 0.50, 0.75, 1),
    na.rm = TRUE
  )
)

pain_labels <- paste0("Q", seq_len(length(pain_breaks) - 1))

patients <- patients %>%
  mutate(
    pain_group = cut(
      pain,
      breaks = pain_breaks,
      labels = pain_labels,
      include.lowest = TRUE
    )
  )

km_pain <- survfit(Surv(days_enrolled, dropout) ~ pain_group, data = patients)

plot(
  km_pain,
  main = "Retention by Baseline Pain Quartile",
  xlab = "Days",
  ylab = "Proportion Enrolled"
)
legend("bottomright", legend = levels(patients$pain_group), lty = 1, bty = "n")
```

## Procedure Groups

```{r km-procedure, fig.height=9}
par(mfrow = c(2, 1))

km_procedure_2group <- survfit(
  Surv(days_enrolled, dropout) ~ procedure_2group,
  data = patients
)
plot(
  km_procedure_2group,
  main = "Retention by Procedure Group — Two Groups",
  xlab = "Days",
  ylab = "Proportion Enrolled"
)
legend("bottomright", legend = levels(patients$procedure_2group), lty = 1, bty = "n")

km_procedure_3group <- survfit(
  Surv(days_enrolled, dropout) ~ procedure_3group,
  data = patients
)
plot(
  km_procedure_3group,
  main = "Retention by Procedure Group — Three Groups",
  xlab = "Days",
  ylab = "Proportion Enrolled"
)
legend("bottomright", legend = levels(patients$procedure_3group), lty = 1, bty = "n")

par(mfrow = c(1, 1))
```

## Diagnosis Groups

```{r km-diagnosis}
km_diagnosis <- survfit(
  Surv(days_enrolled, dropout) ~ diagnosis_2group,
  data = patients
)

plot(
  km_diagnosis,
  main = "Retention by Diagnosis Group",
  xlab = "Days",
  ylab = "Proportion Enrolled"
)
legend("bottomright", legend = levels(patients$diagnosis_2group), lty = 1, bty = "n")
```

# 4. Log-Rank Tests

```{r log-rank-tests}
log_rank_tests <- list(
  mobility = survdiff(Surv(days_enrolled, dropout) ~ mobility, data = patients),
  mood = survdiff(Surv(days_enrolled, dropout) ~ mood, data = patients),
  sleep = survdiff(Surv(days_enrolled, dropout) ~ sleep, data = patients),
  pain_group = survdiff(Surv(days_enrolled, dropout) ~ pain_group, data = patients),
  procedure_2group = survdiff(Surv(days_enrolled, dropout) ~ procedure_2group, data = patients),
  procedure_3group = survdiff(Surv(days_enrolled, dropout) ~ procedure_3group, data = patients),
  diagnosis_2group = survdiff(Surv(days_enrolled, dropout) ~ diagnosis_2group, data = patients)
)

log_rank_summary <- bind_rows(
  lapply(names(log_rank_tests), function(test_name) {
    tibble(
      variable = test_name,
      p_value = glance(log_rank_tests[[test_name]])$p.value
    )
  })
)

kable(log_rank_summary, digits = 4, caption = "Log-Rank Test Summary")
```

# 5. Cox Proportional Hazards Models

## Baseline Clinical Metrics

```{r cox-baseline}
patients <- patients %>%
  mutate(
    pain_centered = pain - mean(pain, na.rm = TRUE),
    pain_centered_squared = pain_centered^2
  )

cox_baseline <- coxph(
  Surv(days_enrolled, dropout) ~
    pain_centered + pain_centered_squared + mobility + mood + sleep,
  data = patients
)

kable(
  tidy(cox_baseline, exponentiate = TRUE, conf.int = TRUE),
  digits = 3,
  caption = "Cox Model — Baseline Clinical Metrics"
)
```

## Procedure Group

```{r cox-procedure}
cox_procedure <- coxph(
  Surv(days_enrolled, dropout) ~
    pain + mobility + mood + sleep + procedure_2group,
  data = patients
)

kable(
  tidy(cox_procedure, exponentiate = TRUE, conf.int = TRUE),
  digits = 3,
  caption = "Cox Model — Procedure Group"
)
```

## Procedure and Diagnosis

```{r cox-procedure-diagnosis}
cox_final <- coxph(
  Surv(days_enrolled, dropout) ~
    pain + mobility + mood + sleep + procedure_2group + diagnosis_2group,
  data = patients
)

kable(
  tidy(cox_final, exponentiate = TRUE, conf.int = TRUE),
  digits = 3,
  caption = "Final Cox Proportional Hazards Model"
)
```

# 6. Procedure × Diagnosis Exploration

```{r procedure-diagnosis-groups}
patients <- patients %>%
  mutate(
    procedure_diagnosis_group = case_when(
      procedure_2group == "Evaluation and Management" & diagnosis_2group == "Musculoskeletal" ~
        "Evaluation and Management + Musculoskeletal",
      procedure_2group == "Other" & diagnosis_2group == "Musculoskeletal" ~
        "Other Procedures + Musculoskeletal",
      procedure_2group == "Evaluation and Management" & diagnosis_2group == "Other" ~
        "Evaluation and Management + Other Diagnoses",
      procedure_2group == "Other" & diagnosis_2group == "Other" ~
        "Other Procedures + Other Diagnoses"
    )
  )

combined_group_counts <- patients %>%
  count(procedure_diagnosis_group, name = "count")

kable(combined_group_counts, caption = "Procedure and Diagnosis Combination Groups")

km_combined <- survfit(
  Surv(days_enrolled, dropout) ~ procedure_diagnosis_group,
  data = patients
)

plot(
  km_combined,
  main = "Retention by Procedure and Diagnosis Group",
  xlab = "Days",
  ylab = "Proportion Enrolled"
)
legend(
  "bottomright",
  legend = levels(factor(patients$procedure_diagnosis_group)),
  lty = 1,
  cex = 0.75,
  bty = "n"
)

combined_log_rank <- survdiff(
  Surv(days_enrolled, dropout) ~ procedure_diagnosis_group,
  data = patients
)

glance(combined_log_rank)
```

# 7. Final Logistic Regression with Sentiment

The sentiment file is created by `03_sentiment_analysis.py`. The cleaned Python script writes a continuous signed `sentiment_score`, so this notebook can use that feature directly instead of reconstructing it from a majority label.

```{r sentiment-summary}
sentiment <- read.csv(sentiment_path, stringsAsFactors = FALSE)

required_sentiment_columns <- c("patient_id", "sentiment_score")
missing_columns <- setdiff(required_sentiment_columns, names(sentiment))

if (length(missing_columns) > 0) {
  stop(
    paste(
      "Sentiment file is missing required columns:",
      paste(missing_columns, collapse = ", ")
    )
  )
}

sentiment_summary <- sentiment %>%
  group_by(patient_id) %>%
  summarise(
    avg_sentiment = mean(sentiment_score, na.rm = TRUE),
    .groups = "drop"
  )

model_data <- patients %>%
  left_join(sentiment_summary, by = "patient_id") %>%
  drop_na(avg_sentiment)
```

A valid nested model comparison requires both models to use the same observations. Therefore, both models below are fit on `model_data` after removing patients without a usable sentiment score.

```{r logistic-models}
logit_without_sentiment <- glm(
  dropout ~ pain + mobility + mood + sleep + procedure_2group + diagnosis_2group,
  data = model_data,
  family = binomial()
)

logit_with_sentiment <- glm(
  dropout ~ pain + mobility + mood + sleep + procedure_2group + diagnosis_2group + avg_sentiment,
  data = model_data,
  family = binomial()
)

kable(
  tidy(logit_with_sentiment, exponentiate = TRUE, conf.int = TRUE),
  digits = 3,
  caption = "Final Logistic Regression — Odds Ratios"
)
```

## Model Comparison

```{r logistic-model-comparison}
likelihood_ratio_test <- anova(
  logit_without_sentiment,
  logit_with_sentiment,
  test = "Chisq"
)

likelihood_ratio_test

AIC(
  Without_Sentiment = logit_without_sentiment,
  With_Sentiment = logit_with_sentiment
)
```

# 8. Dropout Rate by Procedure Group

```{r procedure-dropout-rate}
procedure_dropout <- patients %>%
  group_by(procedure_2group) %>%
  summarise(
    dropout_rate = mean(dropout),
    patients = n(),
    .groups = "drop"
  )

ggplot(procedure_dropout, aes(x = procedure_2group, y = dropout_rate)) +
  geom_col() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Dropout Rate by Procedure Group",
    x = "Procedure Group",
    y = "Dropout Rate"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 15, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```
